---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit the README.Rmd file -->

# featscreen

**featscreen** is an R package providing ready-to-use functions for 
straightforward feature screening in both supervised and unsupervised settings.

**featscreen** is crafted with the overarching goal of simplifying and unifying 
commonly used screening techniques into a single, accessible R package.

Explore the variety of supervised and unsupervised screening techniques 
available in **featscreen** to tailor your feature selection process according 
to your analysis needs. 

The following table provides an overview of the supervised methods currently 
supported in the package:

```{r supervised, echo=FALSE}
# create table
supervised = featscreen::listAvailableScreeningMethods(
  x = 'supervised', 
  show.data.type = TRUE, 
  simplify = TRUE,
  show.category = TRUE
)

# remove
supervised$id = NULL
supervised$multiresponse = NULL

# order
supervised = supervised[order(supervised$category),,drop=F]

# render
knitr::kable(
  x = supervised,
  align = 'crcccc',
  row.names = FALSE
)
```

Continuing our exploration of feature screening capabilities in **featscreen**, 
the following table offers insights into the unsupervised screening methods 
integrated into the package:

```{r unsupervised, echo=FALSE}
# create table
unsupervised = featscreen::listAvailableScreeningMethods(
  x = 'unsupervised', 
  show.data.type = TRUE, 
  simplify = TRUE
)

# remove ids
unsupervised$id = NULL

# render
knitr::kable(
  x = unsupervised,
  align = 'rcccc',
  row.names = FALSE
)
```


## Installation

You can install latest development version from GitHub (requires [devtools](https://github.com/hadley/devtools) package):

```{r install_dev, eval=FALSE}
if (!require("devtools")) {
  install.packages("devtools")
}

devtools::install_github(
  repo = "alebarberis/featscreen", 
  dependencies = TRUE, 
  build_vignettes = FALSE
)
```

## Getting started

If you are just getting started with **featscreen**, we recommend looking at the
[Getting Started](articles/featscreen.html) section of the site.

## Credits
**featscreen** was conceived, developed, and is maintained by Alessandro Barberis ([@alebarberis](https://github.com/alebarberis)).

## Acknowledgements
I would like to express my sincere gratitude to [Prostate Cancer UK](https://prostatecanceruk.org/) for their generous funding, which made it possible for me to develop the first version of this package.
