---
title: "An introduction to `featscreen`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An introduction to `featscreen`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction
**featscreen** is a versatile and user-friendly package designed to simplify the 
process of supervised and unsupervised feature screening in R.

The cornerstone of **featscreen** is the `?screen` function, that can be used 
to effortlessly perform the three screening steps:

1. **Computation of the Statistic**: Harness the power of various statistical 
techniques to compute relevant measures, allowing you to gain insights into the 
relationships and characteristics of your features.
2. **Aggregation of Results for Multi-Response Data**: In scenarios where your 
data involves multiple responses, the results from step 1 are aggregated before 
selection.
3. **Selection of Features to Keep**: Features are retained based on the screening 
results. The package offers a range of common options for feature selection.

Ready to dive in? Let's walk through the steps of using `featscreen` to perform 
your first feature screening analysis.

## Setup
Firstly, we load `featscreen` and other needed packages:

```{r setup}
library(featscreen)
```

## Seed
Now we want to set a seed for the random number generation (RNG). In fact, 
different R sessions have different seeds created from current time and process 
ID by default, and consequently different simulation results. By fixing a seed 
we ensure we will be able to reproduce the results of this vignette. 
We can specify a seed by calling `?set.seed`.


```{r set_seed}
#Set a seed for RNG
set.seed(
  #A seed
  seed = 5381L,                   #a randomly chosen integer value
  #The kind of RNG to use
  kind = "Mersenne-Twister",      #we make explicit the current R default value
  #The kind of Normal generation
  normal.kind = "Inversion"       #we make explicit the current R default value
)
```

## 1. Computation of the Statistic
The initial step in the feature screening process involves the computation of a 
statistic that will be utilized for subsequent feature selection.
 
**featscreen** offers a diverse range of both supervised and unsupervised 
screening techniques.

You can explore the currently supported screening methods by calling the 
`?listAvailableScreeningMethods` function. This function returns a table with at 
least two columns:

* `id`: The unique identifier for each screening method, used in function calls.
* `name`: The name of the screening method.

### Screening Methods
To view the list of currently supported supervised screening methods, use the 
`?listAvailableScreeningMethods` function with the `x` parameter set to `supervised`:


```{r list_supervised_screening_methods}
#list screening methods
screening.methods = listAvailableScreeningMethods(x = 'supervised')

#print in table
knitr::kable(x = screening.methods, align = 'rc')
```


Similarly, we can use the `x` parameter to show the unsupervised methods:

```{r list_unsupervised_screening_methods}
#list screening methods
screening.methods = listAvailableScreeningMethods(x = 'unsupervised')

#print in table
knitr::kable(x = screening.methods, align = 'rc')
```



### Screening Functions
The name of the screening functions can be retrieved by calling `?listAvailableScreeningFunctions`.

Again, you can use the `x` parameter to show only the supervised methods:

```{r list_supervised_screening_functions}
#list screening functions
screening.functions = listAvailableScreeningFunctions(x = 'supervised')

#print in table
knitr::kable(x = screening.functions, align = 'rc')
```


Similarly, you can use the `x` parameter to show only the unsupervised methods:

```{r list_unsupervised_screening_functions}
#list screening functions
screening.functions = listAvailableScreeningFunctions(x = 'unsupervised')

#print in table
knitr::kable(x = screening.functions, align = 'rc')
```



Each function is documented. To learn more about a specific method it is possible to use
the `?` operator. For example, let's check the function `?rowPearsonCor`.

```{r supervised_screening_function_documentation}
#See documentation
?rowPearsonCor
```


From the documentation, we can see that this is a wrapper function that 
internally calls `?row_cor_pearson` from the `matrixTests` package. 

It accepts four arguments in input:

* `x`: A matrix or a data frame.
* `y`: A numerical vector.
* `alternative`: The alternative hypothesis for each row of `x`.
* `conf.level`: The confidence levels of the intervals.


Let's compute the Pearson's correlation for each row of a matrix by using this 
function:

```{r example1}
#Data
x = matrix(rnorm(10 * 20), 10, 20)
y = rnorm(20)

#Compute
rowPearsonCor(x = x, y = y)
```


The function returns a list with two elements:

* `statistic`: A vector containing the values of the correlation coefficient t-test statistic.
* `significance`: A vector containing the p-values of the correlation coefficient t-test.

## 2. Aggregation of Results
**featscreen** provides a straightforward function named `?multiresponse` for the 
aggregation of results from multi-response data.

Let's explore this function:

```{r multiresponse_aggregation_function_documentation}
#See documentation
?multiresponse
```


From the documentation, we learn that the function accepts three arguments:

* `x`: A matrix containing the values to combine.
* `multi`: The aggregation strategy to use. There are five available options:
    + `max`: Keep the maximum value across responses.
    + `min`: Keep the minimum value across responses.
    + `avg`: Average the values.
    + `sum`: Sum up the values.
    + `idx`: Return the column indicated by the argument `idx`.
* `idx`: An optional integer value or string indicating the column of `x` to keep.


## 3. Selection of Features
The final step in the feature screening process involves the selection of 
features to retain based on the computed statistics.

### Selection Methods
The name of the selection methods can be retrieved by calling `?listAvailableSelectionMethods`:

```{r list_selection_methods}
#list selection methods
selection.methods = listAvailableSelectionMethods()

#print in table
knitr::kable(x = selection.methods, align = 'rc')
```

**featscreen** provides five selection strategies:

1. `cutoff`: Features are selected by a cutoff on their values.
2. `rank`: Features are selected based on their ranking.
3. `percentile`: Features are selected by a percentile of the highest values.
4. `fpr`: Features are selected by a cutoff on the false positive rates.
5. `fdr`: Features are selected by a cutoff on the false discovery rates.


### Selection Functions
The name of the selection functions can be retrieved by calling `?listAvailableSelectionFunctions`.

```{r list_selection_functions}
#list selection functions
selection.functions = listAvailableSelectionFunctions()

#print in table
knitr::kable(x = selection.functions, align = 'rc')
```


Each function is documented. To learn more about a specific method it is possible to use
the `?` operator. For example, let's check the function 
`?selectByCutoff`.

```{r selection_function_documentation}
#See documentation
?selectByCutoff
```

From the documentation, we can see that the function accepts three arguments in 
input:

* `x`: A numerical vector.
* `cutoff`: The cutoff to use in the selection.
* `operator`: A string indicating the relational operator to use.

Let's select elements of a vector by using this function.

```{r example2}
#Select by Cutoff
selectByCutoff(
  x = seq_len(10),
  cutoff = 3,
  operator = "<"
)
```

As reported in the documentation, the function returns a logical vector of the 
same length as the input vector, indicating which elements to retain.


## `screen` Function
Rather than using individual screening functions, you can streamline the process 
by using the `?screen` function. This function provides a unified interface for 
various screening methods, incorporating all three steps of the screening process.

```{r screen_function_documentation, eval=FALSE, include=FALSE}
#See documentation
?screen
```

### Parameters
The parameters in input are:

* `x`: Either a matrix or a data frame where rows represent features, and columns represent observations.
* `y`: Numeric vector of data values.
* `g`: (Optional) Vector specifying the group for corresponding elements of `x`.
* `method`: One of the supported screening techniques. See `?listAvailableScreeningMethods`.
* `...`: Additional arguments to the screening function.
* `multi`: Strategy to adopt in case of multi-response data.
* `idx`: (Optional) Integer value or character string indicating the column of `x` to keep.
* `select.by`: The selection method to use.
* `select.args`: A named list containing the arguments to be passed to the selecting function.

### Output
The function returns an object of class `?featscreen`, representing a set of
screened features.

## `featscreen` S3 Class
The `featscreen` class represents a set of screened features. 

A `featscreen` object is a `list` consisting of eight elements:

* `method`: The id of the used screening method.
* `multi`: The id of the used multi-response aggregation method.
* `selection`: The id of the used selection method.
* `summary`: A textual summary of the screening.
* `n`: The dimension of the feature space.
* `features`: The feature names.
* `keep`: The features to keep.
* `ranks`: The feature ranks.

Functions to facilitate access to the data stored in a `featscreen` object are
available:

* `?getScreeningMethodId`: Returns the screening method id.
* `?getMultiresponseAggregationMethodId`: Returns the multi-response aggregation method id.
* `?getSelectionMethodId`: Returns the selection method id.
* `?getSummary`: Returns the textual summary of the screening.
* `?getFeatureDimensionality`: Returns the dimension of the feature space.
* `?getFeatureNames`: Returns the feature names.
* `?getScreenedFeatures`: Returns the features to keep.
* `?getFeatureRanks`: Returns the feature ranks.

Another useful function is `print`:

* `?print.featscreen`: Print a summary of the `featscreen` object.

## Feature Screening
Now let's use the `?screen` function to subset a feature space by using an
unsupervised screening method.

First, let's create the data.

```{r setup_strata}
#Define row/col size
nr = 100
nc = 20

#Data
x = matrix(
  data = sample(x = c(1,2), size = nr*nc, replace = TRUE),
  nrow = nr,
  ncol = nc,
  dimnames = list(
    paste0("feature",seq(nr)),
    paste0("S",seq(nc))
  )
)

#Define grouping variable
g = c(rep("a", nc/2), rep("b", nc/2))

#Force 1st feature to have 40% of missing values
x[1,seq(nc*0.4)] = NA
```

A common unsupervised screening step is to remove a feature if the number of 
missing elements exceeds a defined cutoff. 

### 1. Select Screening Function
If we look at the table returned by `?listAvailableScreeningMethods`, we can see 
that the `id` for feature selection by missing value ratio is `missing.value`. 

Looking at the table returned by `?listAvailableScreeningFunctions`, we can see
that the function corresponding to `missing.value` is `?rowMissingValueRatio`.

From the documentation, we can see that the function accepts two arguments in 
input:

* `x`: matrix or data frame, where rows are features and columns are observations
* `g`: optional grouping vector or factor

If `g` is provided, this function returns a matrix with missing-value ratios for
each group as column vectors.

Let's start building our call.

We can use `missing.value` as the `method` argument in `?screen`. 

To provide an example of multi-response aggregation, we will also provide the
`g` argument through the ellipsis (i.e., the three dots).

### 2. Choose aggregation strategy
As aggregation strategy we choose to keep the maximum value across responses. 
We can do this by providing `max` as the `multi` argument in `?screen`.


### 3. Select Selection Function
Now that we have defined the screening function and the aggregation strategy, we
need to identify the selection method.

Here we use again the function `?selectByCutoff`. 

We want to filter features having more than 50% of missing values, and we can do
this by providing a list containing `cutoff` and `operator` as the `select.args` 
argument in `?screen`.

### 4. Run!
Now let's select the features:

```{r unsupervised_screening_via_screen_function}
#Filter a feature if has more than 50% of missing values
obj = screen(
  x           = x,
  g           = g,
  method      = "missing.value",
  multi       = "max",
  select.by   = "cutoff",
  select.args = list(cutoff = 0.5, operator = "<")
)
```

We can check if the returned object is of class `featscreen`:

```{r check_class_unsupervised_screening}
#Is obj of class `featscreen`?
is.featscreen(obj)
```

Now, we could print a summary:

```{r print_featscreen_obj_unsupervised_screening}
#Print
print(x = obj, show.top = F)
```


## Next Steps
Congratulations! You've successfully completed your first feature screening analysis. 

As you continue to explore the package, refer to the documentation for detailed 
information on each screening method and additional customization options.

Start your **featscreen** journey today and unlock the full potential of efficient 
and effective feature screening in R!
